---
title: "PS2 Group16"
author: "Shailendra Patil, Mingyang Ma, Apurva Gupta"
date: "January 20, 2018"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
```

# Question 1

####Installing NHANES package and ggplot Libraries

```{r}
#install.packages("NHANES")
#install.packages("ggplot2")

```

####Activating NHANES package and ggplot libraries

```{r}
library(NHANES)
library(ggplot2)

```

####Creating a data frame which contains only the people sample aged 18 and up

```{r}
Adults = subset(NHANES, Age >= 18)
head(Adults)

```

####Creating an object of ggplot Library

```{r}
ggobj = ggplot(Adults, aes(x=Adults$Height))
```

####Plotting Histogram for Male and Female Heights

```{r}
ggobj + geom_histogram(binwidth=3)+facet_grid(~Adults$Gender)
```

####Observation
The histogram shows that both male and female heights have approximately same center, spread and distribution.


####Checking Normality by plotting qqplot graphs.

```{r}
 ggplot(Adults, aes(sample=Adults$Height)) + stat_qq() +facet_grid(~Adults$Gender)
```

#####Observation

We can observe that qqplots of male heights and female heights are approximately straight lines. Hence, they are approximately normal.

####Creating scatter plot to find if it is an additive/multiplicative shift.

```{r}
Maleheight=Adults$Height[Adults$Gender=="male"]
Femaleheight=Adults$Height[Adults$Gender=="female"]
qq.df=as.data.frame(qqplot(Maleheight,Femaleheight,plot.it = FALSE))
ggplot(qq.df,aes(x=x,y=y))+geom_point()+geom_abline()
```

#####Observation

We can observe that it is a straight line parallel to reference line. Therefore,  Male and Female heights are in additive shift.

####Quantitative Expression of Male and Female Heights

```{r}
Adults[is.na(Adults$Height),]$Height <-mean(Adults$Height,na.rm =T)
mean_diff = mean(Adults$Height[Adults$Gender=="male"]) - mean(Adults$Height[Adults$Gender=="female"])
mean_diff
```

#####Observation

Quantitatively, We can say that difference in mean heights of male and female is about 14 units.

#Question 2

####Creating an object of ggplot Library

```{r}
ggobj1 = ggplot(Adults, aes(x=Weight))
```

####Plotting Histogram for Male and Female Weights

```{r}
ggobj1 + geom_histogram(binwidth=1)+facet_grid(~Adults$Gender)
```

####Observation
The histogram shows that both male and female weights have approximately same distribution with a slightly shifted center.


####Checking Normality by plotting qqplot graphs.

```{r}
 ggplot(Adults, aes(sample=Adults$Weight)) + stat_qq() +facet_grid(~Adults$Gender)
```

#####Observation

We can observe that qqplots of male Weights and female Weights are curved lines. Hence, we cannot conclude that they are approximately normal.

####Checking Normality by plotting qqplot graphs of log weights.

```{r}
 ggplot(Adults, aes(sample=log(Adults$Weight))) + stat_qq() +facet_grid(~Adults$Gender)
```

#####Observation

We can observe that qqplots of log male and female Weights are appromimately straight lines. Hence, we can conclude that log values are approximately normal.

####Creating scatter plot to find if it is an additive/multiplicative shift.

```{r}
Maleweight=Adults$Weight[Adults$Gender=="male"]
Femaleweight=Adults$Weight[Adults$Gender=="female"]
qq.df=as.data.frame(qqplot(Maleweight,Femaleweight,plot.it = FALSE))
ggplot(qq.df,aes(x=x,y=y))+geom_point()+geom_abline()
```

#####Observation

We can observe that it is approximately a straight line parallel to reference line. Therefore, we can say that Male and Female weights are in additive shift.


####Quantitative Expression of Male and Female Weights

```{r}
Adults[is.na(Adults$Weight),]$Weight <-mean(Adults$Weight,na.rm =T)
mean_diff = mean(Adults$Weight[Adults$Gender=="male"]) - mean(Adults$Weight[Adults$Gender=="female"])
mean_diff
```

#####Observation

Quantitatively, We can say that difference in mean weights of male and female is about 13 units.

# Question 3

####Fitting a linear model to predict Heights from Gender

```{r}
Adult.lm=lm(Height~Gender,data=Adults)
Adult.res=data.frame(Gender=Adults$Gender,residual=residuals(Adult.lm))
Adult.fitted = sort(fitted.values(Adult.lm)) - mean(fitted.values(Adult.lm))
Adult.residuals = sort(residuals(Adult.lm))
```


####Checking normlaity of residuals

```{r}
ggplot(Adult.res, aes(sample = residual)) +
  stat_qq() + facet_wrap(~Gender, ncol=2) 
```

#####Observation

We can observe that residuals are approximately normal.

####Creating a data frame with fiited values and residuals.


```{r}
n = length(Adult.residuals)
f.value = (0.5:(n - 0.5)) / n
Adult.fit = data.frame(f.value, Fitted=Adult.fitted, Residuals=Adult.residuals)
```


#### We will take several variables and gathers them together, so you have more observations but fewer columns.

```{r}
library(tidyr)
adult.fit.long = Adult.fit %>% gather(type, value, Fitted:Residuals)
```


#### Residual fit plot

```{r}
ggplot(adult.fit.long, aes(x=f.value, y=value)) +
  geom_point() + facet_wrap(~type)
```


#####Observation

The fitted values are close together compared to the residuals. While the model may be useful,it only accounts for a fraction of the variation in the data.We can check the results quantitatively.

#### Quantifying the model fit

```{r}
var(Adult.fitted) / var(Adults$Height)
```

#####Observation
the model captures 47% of the variance of the data." The remaining 53% is in the residuals.